Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT

### 2012-08-07 talk/vmil2012/paper.tex
a note
s/PyPy/RPython in some places

Based on the informal observation that guards are among the most common
operations in the traces produced by PyPy's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
the execution state to be able to rebuild it in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of PyPy's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
PyPy's tracing just-in-time compiler, the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of PyPy's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks.
  \item We provide a detailed measurements about the frequency and the
  overhead associated with guards.
  \item We provide a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}
\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for PyPy's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to restore the interpreter state in case of a guard
failure, once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}.\todo{decide if LuaJIT is a footnote or a reference and
fix website citation} Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to PyPy's resume data, to store the information about
how to rebuild the state from a side-exit using the information in the snapshot
and the machine execution state. Pall also acknowledges that snapshot for
guards are associated with a large memory footprint. The solution used in
LuaJIT is to store sparse snapshots, avoiding the creation of snapshots for
every guard to reduce memory pressure. Snapshots are only created for guards
after updates to the global state, after control flow points from the original
program and for guards that are likely to fail. As an outlook Pall mentions the
plans to switch to compressed snapshots to further reduce redundancy.

@@

The deoptimization information of Hotspot is extended
to support correct behaviour
when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
The approach is extremely similar to how RPython's JIT handles virtual objects.
For every object that is not allocated in the code,
the deoptimization information contains a description
of the content of the fields.
When deoptimizing code, these objects are reallocated
and their fields filled with the values
described by the deoptimization information.
The paper does not describe any attempts to store this information compactly.

>>>>

Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
the execution state to be able to rebuild it in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler, the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of RPython's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks.
  \item We provide a detailed measurements about the frequency and the
  overhead associated with guards.
  \item We provide a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}
\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to restore the interpreter state in case of a guard
failure, once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}.\todo{decide if LuaJIT is a footnote or a reference and
fix website citation} Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information about
how to rebuild the state from a side-exit using the information in the snapshot
and the machine execution state. Pall also acknowledges that snapshot for
guards are associated with a large memory footprint. The solution used in
LuaJIT is to store sparse snapshots, avoiding the creation of snapshots for
every guard to reduce memory pressure. Snapshots are only created for guards
after updates to the global state, after control flow points from the original
program and for guards that are likely to fail. As an outlook Pall mentions the
plans to switch to compressed snapshots to further reduce redundancy.

@@

The deoptimization information of Hotspot is extended
to support correct behaviour
when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
The approach is extremely similar to how RPython's JIT handles virtual objects.
For every object that is not allocated in the code,
the deoptimization information contains a description
of the content of the fields.
When deoptimizing code, these objects are reallocated
and their fields filled with the values
described by the deoptimization information.
The paper does not describe any attempts to store this information compactly.
This may not be needed in their approach, because method-based JITs have a lot
fewer deoptimization points than tracing JITs.
